---
title: "BaCoN - Tutorial"
author: "Thomas Rohde"
date: "Feb 22, 2025"
output: html_document
---

This vignette describes how to predict functional buffering between gene pairs using `BaCoN`, using data from the Cancer Dependency Map (DepMap). 


```{r, eval=FALSE, include=FALSE, echo=FALSE}
rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
```

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

## 0. Install BaCoN package

```{r, eval=FALSE, warning=FALSE}
if(!require(devtools)) {install.packages("devtools")}
if (F) {devtools::install_github("billmannlab/BaCoN", force = T)}
```


## 1. Import packages

```{r, warning=FALSE, message=FALSE}
library(BaCoN)
library(data.table)
library(stringr)
library(tidyverse)
theme_set(theme_light())

```


## 2. Data preparation

This script depends on DepMap [gene expression data](https://depmap.org/portal/data_page/?tab=allData&releasename=DepMap%20Public%2024Q4&filename=OmicsExpressionProteinCodingGenesTPMLogp1.csv) as well as [Chronos scores](https://depmap.org/portal/data_page/?tab=allData&releasename=DepMap%20Public%2024Q4&filename=CRISPRGeneEffect.csv), available at the DepMap portal. 

Please download the files and place them in the "data" directory.

- Chronos scores: "CRISPRGeneEffect.csv"
- Gene expression: "OmicsExpressionProteinCodingGenesTPMLogp1.csv"

## 2.1 Import DepMap gene expression and fitness effect data

```{r import input data}

import_depmap <- \(filepath) {
  object <- as.matrix(data.table::fread(filepath), rownames = 1)
  colnames(object) <- stringr::str_split_i(colnames(object), " \\(", 1)
  return(object)
}

ge_complete <- import_depmap(
  file.path("..", "data", "OmicsExpressionProteinCodingGenesTPMLogp1.csv"))

chronos_complete <- import_depmap(
  file.path("..", "data", "CRISPRGeneEffect.csv"))

message(str_c(round(sum(is.na(chronos_complete)) / 
                      length(chronos_complete) * 100, 2), 
              " percent of values are imputed."))

# Impute missing Chronos scores as gene-wise means

for (i in 1:ncol(chronos_complete)) {
  chronos_complete[,i][is.na(chronos_complete[,i])] <- mean(chronos_complete[,i], na.rm = T)
}


```




## 2.2 Subset expression and fitness effect matrix


The gene expression and the Chronos score matrix are reduced to the cell lines that are represented in both datasets. 

```{r}
intersecting_cell_lines <- intersect(rownames(ge_complete), rownames(chronos_complete))

ge_subset <- ge_complete[intersecting_cell_lines,]
chronos_subset <- chronos_complete[intersecting_cell_lines,]

```

The correlations are computed across `r length(intersecting_cell_lines)` interesecting cell lines. 

Conceptually, we expect that biologically meaningful buffering predictions require the buffering partner to be expressed in a sufficient number of cell lines and the knockout of the buffered partner to have a certain impact on cellular fitness. Our experience showed that removing genes with low signal by applying a threshold on the expression level improves buffering predictions. 

This is only one possible strategy. In contrast to correlation scores, BaCoN scores are compared with and thus influenced by all other gene partners of the respective pair. It is recommended to experiment with different matrix sizes. 

To keep runtimes low, we apply very strict thresholds in this vignette. 
In a comprehensive analysis, it is strongly recommended to cover a larger fraction of the genome, by defining less strict thresholds. 

Matrix sizes of ~ 6000-16000 genes on each side of the matrix showed good results in our experiments. The cell lines should be kept at a high number


```{r}
# We restrict the gene space to genes that do not show an expression (log2 TPM+1)
# of > 5 in at least 1000 of the cell lines. 

expression_genes <- names(which(apply(ge_subset >= 5, 2, sum, na.rm = T) >= 1000))

ge_subset <- ge_subset[,expression_genes]

chronos_genes <- names(which(apply(abs(chronos_subset) > 1, 2, sum) >= 30))
chronos_subset <- chronos_subset[,chronos_genes]
```


```{r}
dim(ge_subset)
str(ge_subset)
dim(chronos_subset)
str(chronos_subset)
```

The resulting universe is reduced to `r ncol(ge_subset)` expression genes and `r ncol(chronos_subset)` fitness genes. 

```{r}
prediction_matrices <- array(
  data = NA, 
  dim = c(ncol(ge_subset), ncol(chronos_subset), 4), 
  dimnames = list(colnames(ge_subset), 
                  colnames(chronos_subset), 
                  c("PCC", "BaCoN", "Cholesky_PCC", "Cholesky_BaCoN")))

str(prediction_matrices)
```

## 3.1 Compute PCC correlation matrix:

```{r compute correlation matrix}
prediction_matrices[,,"PCC"] <- cor(ge_subset, chronos_subset, 
                                    use = "pairwise.complete.obs")
```

## 3.2 Compute BaCoN matrix:


BaCoN may behave unexpectedly if the input correlation matrix is formatted in an unexpected way. 
To ensure stable and reliable results, please ensure that the input matrix...

1. is a correlation matrix or composed of comparable values scaling between -1 and 1. BaCoN is currently not designed to work e.g. on negative logarithmic p-values or other association indices. 

2. does not contain NAs. This should not cause problems in cases of single NA values, but entire NA rows or columns should be removed from the matrix. 

3. should not have duplicated row- or column names. Each gene should be represented in one row and/or column only. 

```{r compute BaCoN matrix, message=FALSE}
prediction_matrices[,,"BaCoN"] <- BaCoN(prediction_matrices[,,"PCC"])
```




## 4. Cholesky whitening + PCC + BaCoN

Next, we combine `BaCoN` with the pre-hoc covariance normalization method Cholesky whitening, as proposed in our manuscript. 
We use the `whitening` package and its `whiten` function. 


```{r}
if (F) {install.packages(whitening)}
library(whitening)
```

We whiten both input matrices independently. It is important that the cell lines of both objects are identical and in the "columns" dimensions of the object. Therefore, we transpose the objects before and retranspose them after the whitening. 


```{r, message=FALSE}
whitened_ge_subset <- t(whiten(t(ge_subset), method = "Cholesky"))
whitened_chronos_subset <- t(whiten(t(chronos_subset), method = "Cholesky"))

prediction_matrices[,,"Cholesky_PCC"] <- cor(
  whitened_ge_subset, 
  whitened_chronos_subset, use = "pairwise.complete.obs")

prediction_matrices[,,"Cholesky_BaCoN"] <- BaCoN(prediction_matrices[,,"Cholesky_PCC"])

```


## 5. Collect top 100 predictions:

We use the lowest BaCoN score of the top 100 predictions as cutoff. The high density of paralog pairs is clearly visible. 
This can lead to more than 100 pairs in the prediction set, as the top BaCoN-scored pairs are often tied. We use the PCC score to deal with the tied pairs. 

```{r collect predictions}
predictions <- data.table(data.frame(
  expression_gene = rep(rownames(prediction_matrices), 
                        ncol(prediction_matrices)), 
  chronos_gene = rep(colnames(prediction_matrices), 
                     each = nrow(prediction_matrices)), 
  sapply(dimnames(prediction_matrices)[[3]], 
         \(.) as.vector(prediction_matrices[,,.]), simplify = F)))

predictions[order(BaCoN, PCC, decreasing = T), BaCoN_rank := .I]
predictions[order(Cholesky_BaCoN, Cholesky_PCC, decreasing = T), Cholesky_rank := .I]

predictions <- predictions[order(BaCoN, PCC, decreasing = T)]

str(predictions)
```


```{r}
predictions[1:25, .(expression_gene, chronos_gene, PCC, BaCoN, Cholesky_BaCoN)]
```




## 6. Show score distribution and predictions

```{r scatter plot, echo=FALSE, out.width="100%"}
get_density <- \(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

plot_data <- predictions[
  sample(1:.N, 20000), 
  .(PCC, 
    BaCoN, 
    Cholesky_PCC, 
    Cholesky_BaCoN, 
    density = get_density(PCC, BaCoN, n = 500), 
    cholesky_density = get_density(Cholesky_PCC, Cholesky_BaCoN, n = 500))]

ggplot(plot_data, aes(PCC, BaCoN)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_point(aes(color = density), shape = ".") + 
  scale_color_gradientn(colours = c("grey", "navy", "red", "yellow")) + 
  geom_point(data = predictions[BaCoN_rank <= 100], color = "seagreen") + 
  scale_x_continuous(limits = c(-0.6, 0.6)) + 
  labs(title = "BaCoN vs PCC scores", 
       subtitle = str_c("Heatscatter plots of BaCoN ~ PCC coordinates of gene pairs.\n", 
                        "Gene pars with the top 100 BaCoN scores are colored green.")) + 
  theme(legend.position = "bottom")


ggplot(plot_data, aes(Cholesky_PCC, Cholesky_BaCoN)) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_point(aes(color = cholesky_density), shape = ".") + 
  scale_color_gradientn(colours = c("grey", "navy", "red", "yellow")) + 
  geom_point(data = predictions[Cholesky_rank <= 100], color = "seagreen") + 
  scale_x_continuous(limits = c(-0.6, 0.6)) + 
  labs(title = "BaCoN vs PCC scores", 
       subtitle = str_c("Heatscatter plots of Cholesky+BaCoN ~ Cholesky+PCC coordinates of gene pairs.\n", 
                        "Gene pars with the top 100 Cholesky+BaCoN scores are colored green.")) + 
  theme(legend.position = "bottom")


```

# Summary

The exponentially increasing runtime of `BaCoN` is due to the greater number of required computations, 
but also the handling of very large objects. 
The `data.table` environment is designed to optimize speed and RAM usage when dealing with large datasets. 
The function is designed to minimize runtime increase caused by data handling, as well as RAM usage. 

Future versions of `BaCoN` might offer further runtime improvements. 


